---
title: "COPDGene/DECAMP Radiologic Clustering Project"
author: "Samuel Ash"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE,error=FALSE,error=FALSE,warning=FALSE,echo=FALSE)
```

# Overview
This document describes the initial results of the imaging clusters from COPDGene based on prior DECAMP clustering.  This includes the baseline clinical characteristics and biomarkers as well as longitudinal analyses including change in FEV1, change in 6WMD, mortality, and exacerbation rates.

# Load Libraries and Data

## Libraries

```{r load_libs}
library(devtools)
library(tidyverse)
library(janitor)
library(cowplot)
library(survival)
library(survminer)
library(readxl)
library(ggfortify)
library(ggpubr)
library(rmarkdown)
```

## Data
COPDGene datasets used include:
1) Baseline and longitudinal clinical dataset from October 9, 2017
2) Mortality dataset from December 2016
3) LFU dataset from July 2017
4) LH dataset from September 21, 2017 (using third set of LH training data)
5) RBM dataset
6) body composition dataset using manual measurements

```{r data_import}
## 02 Initial Import and Cleaning of COPDGene Data ##
## edited 2018_08_08

#### general COPDGene database ####
copd <- read_delim("../data/raw_data/COPDGene_P1P2_All_Visit_09OCT17.txt", 
                   "\t", escape_double = FALSE, trim_ws = TRUE)
# clean variable names and remove columns and rows with all NA
copd_clean <- clean_names(copd) %>% 
  rename(dlco_pp=dlc_opp, 
         tlc_pp_race_adjusted=tl_cpp_race_adjusted,
         fvcpp_utah=fv_cpp_utah,
         dlco_raw=d_lco_raw) %>% 
  remove_empty()
## calculate change variables ##
# remove subset of variables that we are interested in their change over time
copd_change = copd_clean %>% select(sid,visitnum,height_cm,weight_kg,
                                     distwalked,sgrq_score_total,dlco_pp, dlco_raw,
                                     tlc_pp_race_adjusted,fev1pp_utah,fvcpp_utah,
                                     fev1_fvc_utah,pct_emph_thirona,tlc_thirona,
                                     pct_gas_trap_thirona,awt_seg_thirona,
                                     fev1_utah,fvc_utah)
# spread to wide format
copd_change_wide = copd_change %>% 
  tidyr::gather_(key="variable", value="value", 
                 c("height_cm",
                   "weight_kg",
                   "distwalked",
                   "sgrq_score_total",
                   "dlco_pp", 
                   "dlco_raw",
                   "tlc_pp_race_adjusted",
                   "fev1pp_utah",
                   "fvcpp_utah",
                   "fev1_fvc_utah",
                   "pct_emph_thirona",
                   "tlc_thirona",
                   "pct_gas_trap_thirona",
                   "awt_seg_thirona",
                   "fev1_utah",
                   "fvc_utah"
                   )) %>% 
  tidyr::unite("time_by_variable", variable, visitnum, remove=T) %>%  
  tidyr::spread(key=time_by_variable, value=value)                
# calculate change between two variables
copd_delta = copd_change_wide %>% 
  mutate(awt_seg_thirona_delta = awt_seg_thirona_2 - awt_seg_thirona_1,
         distwalked_delta = distwalked_2 - distwalked_1,
         dlco_pp_delta = dlco_pp_2 - dlco_pp_1,
         dlco_raw_delta = dlco_pp_2 - dlco_raw_1,
         fev1_fvc_utah_delta = fev1_fvc_utah_2 - fev1_fvc_utah_1,
         fev1_utah_delta = fev1_utah_2 - fev1_utah_1,
         fev1pp_utah_delta = fev1pp_utah_2 - fev1pp_utah_1,
         fvc_utah_delta = fvc_utah_2 - fvc_utah_1,
         fvcpp_utah_delta = fvcpp_utah_2 - fvcpp_utah_1,
         height_cm_delta = height_cm_2 - height_cm_1,
         pct_emph_thirona_delta = pct_emph_thirona_2 - pct_emph_thirona_1,
         pct_gas_trap_thirona_delta = pct_gas_trap_thirona_2 - pct_gas_trap_thirona_1,
         sgrq_score_total_delta = sgrq_score_total_2 - sgrq_score_total_1,
         tlc_pp_race_adjusted_delta = tlc_pp_race_adjusted_2 - tlc_pp_race_adjusted_1,
         tlc_thirona_delta = tlc_thirona_2 - tlc_thirona_1,
         weight_kg_delta = weight_kg_2 - weight_kg_1) %>% 
  select(sid,awt_seg_thirona_delta,distwalked_delta,dlco_pp_delta,
         fev1_fvc_utah_delta,fev1_utah_delta,fev1pp_utah_delta,fvc_utah_delta,fvcpp_utah_delta,
         height_cm_delta,pct_emph_thirona_delta,pct_gas_trap_thirona_delta,sgrq_score_total_delta,
         tlc_pp_race_adjusted_delta,weight_kg_delta)
# extract time between visits and copdgene calculated change variables 
copd_time = copd_clean %>% select(sid,days_ct1_ct2,days_phase1_phase2,
                                   change_tlc_thirona, change_pct_emph_thirona, change_pct_gas_trap_thirona,
                                   change_p1_p2_fev1_ml,change_p1_p2_fev1pp) %>% 
  filter(days_phase1_phase2>0)
# remove visit 2 data and finalgold -2 from base dataset 
copd_base = copd_clean %>% filter(visitnum==1) %>% filter(final_gold > -2) %>% select(-visitnum) 
# merge delta variable data with baseline data
copd_pre = inner_join(copd_base,copd_delta)
# add time variables and remove variables that are all NA
copd_base_full = left_join(copd_pre,copd_time,by="sid") %>% remove_empty()
# add prior exacerbation binary variable
copd_base_full = copd_base_full %>% mutate(priorexacerbation = ifelse(exacerbation_frequency > 0, 1, 0))

### mortality and LFU dataset addition to baseline dataset ###
# import datasets
copd_mort <- read_delim("../data/raw_data/COPDGene_Mortality_Surv_2016dec.txt", 
                                                          "\t", escape_double = FALSE, col_types = cols(DOD_merged = col_date(format = "%m/%d/%Y"), 
                                                                                                        Phase1_date = col_date(format = "%m/%d/%Y"), 
                                                                                                        date_SSDI_confirmed = col_date(format = "%m/%d/%Y"), 
                                                                                                        date_last_SSDI_search = col_date(format = "%m/%d/%Y"), 
                                                                                                        days_followed = col_number(), last_LFU_survey_date = col_date(format = "%m/%d/%Y")), 
                                                          trim_ws = TRUE)
copd_lfu <- read_delim("../data/raw_data/LFU_SidLevel_w_Comorbid_31JUL17.txt", 
                       "\t", escape_double = FALSE, col_types = cols( 
                                                                     reportedSmokingStatusString = col_skip()), 
                       trim_ws = TRUE) 
# merge datasets
copd_clinical_full_1 = left_join(copd_base_full,copd_mort)
copd_clinical_full = left_join(copd_clinical_full_1,copd_lfu) %>% remove_empty()

### local histogram data using third training set ###
copd_lh = read_excel("../data/raw_data/LHsubtypes_2016_09_21_v2.xlsx") %>% clean_names()
copd_pre1 = left_join(copd_clinical_full,copd_lh) %>% remove_empty()

### body composition from manual measurements ###
copd_body <- Body_Composition_COPDGene <- read_csv("../data/raw_data/Body Composition COPDGene.csv")
copd_pre2 <- left_join(copd_pre1,copd_body) %>% remove_empty()

### biomarker data from Russ Bowler (RBM only) ###
copd_bio <- read_csv("../data/raw_data/COPDGene_Bowler_RBM_raw.csv") %>% clean_names() %>% remove_empty()
copd_full <- left_join(copd_pre2,copd_bio)

#############additional data manipulation################
# rename gender
copd_full = copd_full %>% mutate(sex = ifelse(gender==2,"female","male"))
# normalize imaging variables
copd_full = copd_full %>% mutate(percent_ild_z = percent_ild, percent_emphysema_z = percent_emphysema, awt_seg_thirona_z = awt_seg_thirona, PMA_z = PMA) %>% 
  mutate_at(scale,.vars = vars(percent_ild_z, percent_emphysema_z, awt_seg_thirona_z, PMA_z))

### dataset with no missing imaging data for clustering ###
copd_full_imaging = copd_full %>% filter(!is.na(percent_ild), !is.na(percent_emphysema), !is.na(awt_seg_thirona), !is.na(PMA))
```


# Analysis
## Clustering
Using k means (k=3 based on DECAMP).  Note that the clusters are relabeled to correspond to the DECAMP clusters based on the local histogram characteristics.

```{r cluster}
### image clustering based on emphysema, ILD, PMA and AWT ###
# note use of ggfortify as described: https://cran.r-project.org/web/packages/ggfortify/vignettes/plot_pca.html

# initial clustering based on imaging features - see https://www.datacamp.com/community/tutorials/k-means-clustering-r
set.seed(9)
copd_cluster_pre = copd_full_imaging %>% select(sid,percent_ild_z,percent_emphysema_z,PMA_z,awt_seg_thirona_z) # dataset with just SID and normalized imaging variables
km = kmeans(copd_cluster_pre[,2:5],3)
copd_full_imaging$cluster <- as.factor(km$cluster) # add initial cluster label to dataset of people with complete imaging data
copd_full_imaging = copd_full_imaging %>% mutate(cluster_decamp = ifelse(cluster=="1","B", ifelse(cluster=="2","C","A")))
```

### Cluster appearance:
```{r cluster_plot,fig.width=10,fig.height=7}
autoplot(prcomp(copd_cluster_pre[,2:5]), data=copd_full_imaging, colour='cluster_decamp')
```

### Comparison of clusters based on imaging characteristcs:
```{r cluster_imaging,fig.width=10,fig.height=10}
cluster_comparisons <- list( c("A", "B"), c("B", "C"), c("A", "C") )

emph_cluster_plot =  ggplot(copd_full_imaging, aes(cluster_decamp,percent_emphysema)) + 
  geom_boxplot(fill="gray") + 
  ylab("Percentage Emphysema") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y = 80, method="anova")     # Add global p-value
ild_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,percent_ild)) + 
  geom_boxplot(fill="gray") + 
  ylab("Percentage Interstitial Changes") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y = 32, method="anova")     # Add global p-value
awt_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,awt_seg_thirona)) + 
  geom_boxplot(fill="gray") + 
  ylab("Airway Wall Thickness") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y = 2.2, method="anova")     # Add global p-value
pma_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,PMA)) + 
  geom_boxplot(fill="gray") + 
  ylab("PMA") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y = 120, method="anova")     # Add global p-value


plot_grid(emph_cluster_plot,ild_cluster_plot,awt_cluster_plot,pma_cluster_plot, ncol=2)
```

### Mortality of the clusters
```{r mortality,fig.width=10,fig.height=7}
# kaplan meier curve for mortality
km_fit <- survfit(Surv(days_followed, vital_status) ~ cluster_decamp, data=copd_full_imaging)
ggsurvplot(km_fit, data=copd_full_imaging, 
           risk.table=TRUE,
           pval = TRUE,
           break.time.by = 500,
           ggtheme = theme_minimal(),
           risk.table.y.text.col = T,
           risk.table.y.text = FALSE,
           legend.title = "Clusters",
           legend.labs = c("A","B","C"))
```


### Comparison of clusters based on lung function, exercise capacity and quality of life
1) Baseline Values
```{r clinical_baseline,fig.width=10,fig.height=10}

fev1_cluster_plot =  ggplot(copd_full_imaging, aes(cluster_decamp,fev1pp_utah)) + 
  geom_boxplot(fill="gray") + 
  ylab("FEV1") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y = 220, method="anova")     # Add global p-value
fvc_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,fvcpp_utah)) + 
  geom_boxplot(fill="gray") + 
  ylab("FVC") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y = 220, method="anova")     # Add global p-value
sixmin_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,distwalked)) + 
  geom_boxplot(fill="gray") + 
  ylab("6 Minute Walk Distance") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y = 3800, method="anova")     # Add global p-value
sgrq_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,sgrq_score_total)) + 
  geom_boxplot(fill="gray") + 
  ylab("SGRQ") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y = 150, method="anova")     # Add global p-value

plot_grid(fev1_cluster_plot,fvc_cluster_plot,sixmin_cluster_plot,sgrq_cluster_plot, ncol=2)
```

2) Progression over 5 Years of Follow Up
```{r clinical_progression,fig.width=10,fig.height=5}
fev1prog_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,fev1_utah_delta)) + 
  geom_boxplot(fill="gray") + 
  ylab("FEV1 Change") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y = 5, method="anova")     # Add global p-value
distwalked_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,distwalked_delta)) + 
  geom_boxplot(fill="gray") + 
  ylab("6MWD Change") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y = 2500, method="anova")     # Add global p-value

plot_grid(fev1prog_cluster_plot,distwalked_cluster_plot, ncol=2)
```

### Exacerbation Rate by Cluster
Note that this comparison is only in those who reported at least one exacerbation during follow up
```{r exacerbation,fig.width=10,fig.height=5}
copd_full_imaging <- copd_full_imaging %>% 
  mutate(exacerbation_rate = Total_Exacerbations/Years_Followed,
         sev_exacerbation_rate = Total_Severe_Exacer/Years_Followed)

exacerbation_cluster_plot =  ggplot(subset(copd_full_imaging, priorexacerbation == 1), aes(cluster_decamp,exacerbation_rate)) + 
  geom_boxplot(fill="gray") + 
  ylab("Exacerbation Rate") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y = 25)     # Add global p-value

sev_exacerbation_cluster_plot =  ggplot(subset(copd_full_imaging, sev_exacerbation_rate > 0), aes(cluster_decamp,sev_exacerbation_rate)) + 
  geom_boxplot(fill="gray") + 
  ylab("Severe Exacerbation Rate") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y = 20)     # Add global p-value

plot_grid(exacerbation_cluster_plot, sev_exacerbation_cluster_plot)
```

### Selected Biomarkers by Cluster
```{r biomarkers,fig.width=15,fig.height=30}
crp_cluster_plot =  ggplot(copd_full_imaging, aes(cluster_decamp,crp)) + 
  geom_boxplot(fill="gray") + 
  ylab("CRP") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y = 75, method="anova")     # Add global p-value
tgf_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,tgfb1_lap)) + 
  geom_boxplot(fill="gray") + 
  ylab("TGFb LAP1") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y = 25, method="anova")     # Add global p-value
mmp9_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,mmp9)) + 
  geom_boxplot(fill="gray") + 
  ylab("MMP9") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y = 3500, method="anova")     # Add global p-value
ager_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,ager)) + 
  geom_boxplot(fill="gray") + 
  ylab("AGER") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y=30, method="anova")     # Add global p-value
tnfrsf1b_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,tnfrsf1b)) + 
  geom_boxplot(fill="gray") + 
  ylab("TNFFRSF1B") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y=35,method="anova")     # Add global p-value
icam1_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,icam1)) + 
  geom_boxplot(fill="gray") + 
  ylab("ICAM1") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y=575,method="anova")     # Add global p-value
serpine1_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,serpine1)) + 
  geom_boxplot(fill="gray") + 
  ylab("SERPINE1") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y=250,method="anova")     # Add global p-value
vegfa_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,vegfa)) + 
  geom_boxplot(fill="gray") + 
  ylab("VEGF Alpha") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y=600,method="anova")     # Add global p-value
il1a_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,il1a)) + 
  geom_boxplot(fill="gray") + 
  ylab("IL1 alpha") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y=0.005,method="anova")     # Add global p-value
ccl20_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,ccl20)) + 
  geom_boxplot(fill="gray") + 
  ylab("CCL20") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y=2000,method="anova")     # Add global p-value
il1b_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,il1b)) + 
  geom_boxplot(fill="gray") + 
  ylab("IL1 beta") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(label.y=5.15,method="anova")     # Add global p-value
il1rn_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,il1rn)) + 
  geom_boxplot(fill="gray") + 
  ylab("IL1 RN") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(method="anova")     # Add global p-value
cxcl5_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,cxcl5)) + 
  geom_boxplot(fill="gray") + 
  ylab("CXCL5") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(method="anova")     # Add global p-value
olr1_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,olr1)) + 
  geom_boxplot(fill="gray") + 
  ylab("OLR1") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(method="anova")     # Add global p-value
serpina3_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,serpina3)) + 
  geom_boxplot(fill="gray") + 
  ylab("SERPINA3") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(method="anova")     # Add global p-value
tnfrsf10c_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,tnfrsf10c)) + 
  geom_boxplot(fill="gray") + 
  ylab("tnfrsf10c") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(method="anova")     # Add global p-value
il8_cluster_plot = ggplot(copd_full_imaging, aes(cluster_decamp,il8)) + 
  geom_boxplot(fill="gray") + 
  ylab("IL8") +
  xlab("Cluster") +
  stat_compare_means(comparisons = cluster_comparisons) + # Add pairwise comparisons p-value
  stat_compare_means(method="anova")     # Add global p-value

plot_grid(crp_cluster_plot,tgf_cluster_plot,mmp9_cluster_plot,ager_cluster_plot, 
          tnfrsf1b_cluster_plot,icam1_cluster_plot,serpine1_cluster_plot,vegfa_cluster_plot,
          il1a_cluster_plot,ccl20_cluster_plot,il1b_cluster_plot,il1rn_cluster_plot,
          cxcl5_cluster_plot,olr1_cluster_plot,serpina3_cluster_plot,tnfrsf10c_cluster_plot,il8_cluster_plot,ncol=3)
```



